\chapter{Problem Setup}
\label{sec:Chapter1} \index{Chapter1}

In this thesis, we consider a standard \textbf{unconstrained minimization} problem in the context of linear regression. Let
\[
  \{(x^{(1)}, y^{(1)}), (x^{(2)}, y^{(2)}), \dots, (x^{(n)}, y^{(n)})\}
\]
be a set of $n$ training examples, where each $x^{(i)} \in \mathbb{R}^d$ is a feature vector, and each $y^{(i)} \in \mathbb{R}$ is the corresponding target. We assume there exists an unknown parameter vector $w^* \in \mathbb{R}^d$ such that 
\[
  y^{(i)} \;=\; \langle x^{(i)}, \, w^* \rangle 
  \quad \text{(possibly up to additive noise).}
\]

Our goal is to recover or approximate $w^*$ by minimizing the empirical risk:
\begin{equation}
  \label{eq:empirical-risk}
  R_{w^*}(w) \;=\; \frac{1}{2n} \Bigl\|\mathbf{X}^{\top} w \;-\; \mathbf{X}^{\top} w^*\Bigr\|^2,
\end{equation}
where $\mathbf{X}$ is the $d \times n$ matrix whose columns are the vectors $x^{(i)}$. 
In classical optimization terms, we aim to solve:
\[
  \min_{w \,\in\, \mathbb{R}^d}\; R_{w^*}(w).
\]
When $\mathbf{X}$ has full column rank (or $x^{(i)}$ are sampled from a non-degenerate distribution), 
the function $R_{w^*}(w)$ is convex and differentiable.

\subsection{Iterative First-Order Methods}

Many first-order optimization methods solve
\[
  \min_{w}\; f(w)
\]
by repeatedly updating $w$ using the gradient $\nabla f(w)$. A \emph{linear first-order method} (LFOM) 
can be written generally as
\begin{equation}
  \label{eq:LFOM}
  w_{k+1} \;=\; w_0 \;+\; \sum_{i=0}^{k}\Gamma_k^i\,\nabla f\bigl(w_i\bigr),
\end{equation}
where each $\Gamma_k^i$ is a (diagonal) matrix that scales the $i$-th gradient. This broad family includes, for example, gradient descent, momentum methods, and conjugate gradient descent for quadratic objectives.

\subsection{Memory-Augmented Transformers (Memformers)}

Recent studies show that Transformers \textbf{ memory-augmented} (also known as \emph{Memformers}) can \emph{implement} such iterative updates \emph{in-context} within their forward pass. Specifically, a Memformer can store and reuse intermediate computations (akin to saving previous gradients or partial solutions) via attention-based memory registers. Consequently, one can view the forward computation of a Memformer as performing gradient-based updates on \eqref{eq:empirical-risk}, effectively solving the linear regression problem. 

Formally, if we denote the iteration at step $k$ by $w_k$, Memformers learn to realize updates of the form:
\[
  w_{k+1} \;=\; \text{update}\bigl(w_k,\;\nabla R_{w^*}(w_k),\;\text{memory}\bigr),
\]
where the \emph{memory} captures relevant quantities like past gradients. This parallels the LFOM expression in \eqref{eq:LFOM} but is carried out through a sequence of transformer layers and attention blocks.

\subsection{Unconstrained Quadratic Example}

In linear regression,
\[   R_{w^*}(w) \;=\; \frac{1}{2n}   \Bigl\|\mathbf{X}^\top w \;-\; \mathbf{X}^\top w^*\Bigr\|^2, \],
which is a convex quadratic function. Popular methods like gradient descent, momentum, or conjugate gradient all iteratively minimize this function. 
Memformers replicate these methods by encoding partial iterates of the optimization trajectory into the attention layers.

Thus, the learning goal optimization $w$ that minimizes \eqref{eq:empirical-risk} can be viewed as
\[
  \min_{w \,\in\, \mathbb{R}^d} \quad 
  \frac{1}{2n}\|\mathbf{X}^\top w - \mathbf{X}^\top w^*\|^2,
\]
and is solved \emph{in-context} by the Memformer’s learned parameters and attention-based memory mechanism.


\subsection{Lemma 1 \cite{ahn2024transformers}.}
We consider an $L$-layer linear Transformer with parameter matrices 
\[
P_\ell = 
\begin{bmatrix}
\mathbf{B}_\ell = 0_{d \times d} & 0 \\
0 & 1
\end{bmatrix},
\quad
Q_\ell = 
-\begin{bmatrix}
\mathbf{A}_\ell & 0 \\
0 & 0
\end{bmatrix}, 
\quad \mathbf{A}_\ell, \mathbf{B}_\ell \in \mathbb{R}^{d \times d}.
\]
Let $y_\ell^{(n+1)}$ denote the $(d+1,\,n+1)$-entry of the $\ell$-th layer output matrix~$\mathbf{Z}_\ell$, i.e., 
\(
   y_\ell^{(n+1)} = [\mathbf{Z}_\ell]_{(d+1),\,(n+1)}
\)
for $\ell = 1, \dots, L$. Then
\[
   y_\ell^{(n+1)} \;=\; -\,\bigl\langle \mathbf{x}^{(n+1)}, \,\mathbf{w}_\ell^{\mathrm{gd}}\bigr\rangle,
\]
where the sequence $\{\mathbf{w}_\ell^{\mathrm{gd}}\}$ is initialized by $\mathbf{w}_{0}^{\mathrm{gd}} = 0$ and, for $\ell = 1, \dots, L - 1,$ follows
\[
   \mathbf{w}_{\ell+1}^{\mathrm{gd}}
   \;=\;
   \mathbf{w}_{\ell}^{\mathrm{gd}}
   \;-\;
   \mathbf{A}_\ell 
   \,\nabla R_{\mathbf{w}^*}\bigl(\mathbf{w}_{\ell}^{\mathrm{gd}}\bigr),
\]
with the empirical least-squares loss 
\[
   R_{\mathbf{w}^*}(w) 
   \;:=\; 
   \frac{1}{2n}\,\bigl\|\mathbf{X}^\top w - \mathbf{X}^\top w^*\bigr\|^2
   \;=\;
   \frac{1}{2n}\,
   (w - w^*)^\top \,\mathbf{X}\,\mathbf{X}^\top \,(w - w^*).
\]
